Assessing the Firing of Ceramic Materials: a Seriation-Based Approach

### 2023-05-19 analysis/paper/paper.qmd
Draft results

This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data, which aims to provide an alternative to the EFT framework. This method is based on the principle of parsimony with two main objectives: (1) to avoid the use of external references (whether experimental or ethnographic data); (2) to build the simplest possible model by limiting the number of hypotheses involved.

XRD analysis of archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing. Quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content. The estimation of quantitative mineralogical compositions, by the RIR method [@hubbard1988] or the Rietvelt refinement [@rietveld1969], is a very difficult, high skilled exercise, assuming a thorough knowledge of the material (and expensive databases).

In this study, we show that it is possible to successfully use correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials. The properties of correspondence analysis and its links with matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers. Thus, calibration data sets can be produced on the only basis of archaeological evidence to assess ceramic firing practices across a cultural area, avoiding the use of laboratory references.

@@

@@

Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows to describe the statistical relationships that can exist between individuals and variables. Unlike PCA, CA analyses the differences between relative values and it considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space). CA studies the inertia of the scatterplot (i.e. the weighted sum of the $\chi2$ distances of each point to the centroid) to create orthogonal components, so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the component, and so on until the last dimension. Instead of maximizing variance, CA maximizes the correspondence between individuals and variables.

@@

pkg <- vapply(
  X = dep[dep != "base"], 
  FUN = function(x) {
    sprintf("`%s` %s [@R-%s]", x, packageVersion(x), x)
  }, 
  FUN.VALUE = character(1)
)
```

@@

>>>>

This study therefore proposes a new method, based on multivariate analysis of powder X-ray diffraction (XRD) data. It aims to provide an alternative to the EFT framework. This method is based on the principle of parsimony with two main objectives: (1) to avoid the use of external references (whether experimental or ethnographic data); (2) to build the simplest possible model by limiting the number of hypotheses involved.

XRD analysis of archaeological ceramics allows the identification of the main minerals and special attention is usually paid to the presence (or absence) of mineralogical phases formed during firing. However, quantitative analysis of XRD data faces major difficulties, as archaeological ceramics are multiphase systems, often poorly crystallised, with amorphous content. The estimation of quantitative mineralogical compositions, by the RIR method [@hubbard1988] or the Rietvelt refinement [@rietveld1969], is a very difficult, high skilled exercise, assuming a thorough knowledge of the material (and expensive databases).

In this study, we show that it is possible to successfully use correspondence analysis (CA) of X-ray diffractograms to assess the firing of ceramic materials. The properties of correspondence analysis and its application for matrix seriation enable to characterise the firing of a ceramic assemblage and to identify the outliers. Thus, calibration data sets can be produced on the only basis of archaeological evidence to assess ceramic firing practices across a cultural area, avoiding the use of laboratory references.

@@

Finally, all the diffractograms were combined into a matrix, where each row corresponds to a sample and each column to a $2\theta$ position, so that each cell contains the corresponding X-ray photon count.

@@

Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows to describe the statistical relationships that can exist between individuals and variables. Unlike PCA, CA analyses the differences between relative values and it considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space). CA studies the inertia (i.e. the weighted sum of the $\chi2$ distances of each point to the centroid) to create orthogonal components, so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the component, and so on until the last dimension. CA maximizes the correspondence between individuals and variables, instead of maximizing the amount of variance explained by its reduced-space.

CA is an effective method for the seriation of archaeological assemblages (REF). The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation. The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time. CA is not limited to chronological modeling, but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e. any aspect which is expected to be related to the observed composition (be it temporal or environmental).

@@

pkg <- vapply(
  X = dep[dep != "base" & dep != "here"], 
  FUN = function(x) {
    sprintf("`%s` %s [@R-%s]", x, packageVersion(x), x)
  }, 
  FUN.VALUE = character(1)
)
```

@@

```{r}
#| label: xrd-data
## Read XRD data 
## (see analysis/supplementary-materials/01-data.R)
xrd_counts <- here::here("analysis/data/derived_data/xrd_clean.csv") |> 
  read.table(sep = ",", dec = ".", header = TRUE)

## Remove the first column (2 theta positions)
theta_scale  <- xrd_counts[, 1]
xrd_counts <- xrd_counts[, -1]

## Transpose (one sample per line)
xrd_counts <- xrd_counts |> t() |> as.data.frame()
colnames(xrd_counts) <- paste0("T", seq_len(ncol(xrd_counts)))

## Get index of XRD columns
xrd_col <- which(grepl(pattern = "T", colnames(xrd_counts)))

## Read mineral data
minerals <- here::here("analysis/data/derived_data/minerals.csv") |>
  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8") |> 
  subset(symbole %in% c("An", "Cal", "Gh", "Ilt", "Qz", "Di"))
# minerals <- minerals[!duplicated(minerals$symbole), ]
```

```{r}
#| label: ca-moreno
## Mas de Moreno: ceramics + unfired sherds
sup_row_id <- match(corpus$sample[corpus$site != "Mas de Moreno"], rownames(xrd_counts))
xrd_moreno_ca <- dimensio::ca(object = xrd_counts[, xrd_col], sup_row = sup_row_id)

## Get CA results
xrd_moreno_eig <- dimensio::get_eigenvalues(xrd_moreno_ca)
xrd_moreno_row_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 1)
xrd_moreno_col_coord <- dimensio::get_coordinates(xrd_moreno_ca, margin = 2)
xrd_moreno_col_contrib <- dimensio::get_contributions(xrd_moreno_ca, margin = 2)
```

The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca).

```{r}
#| label: fig-plot-ca
#| fig-cap: Scree plot of inertia (top left; only the first ten components are displayed).
#| fig-width: 10
#| fig-height: 10
## Plot eigenvalues
par(mar = c(4, 4, 1, 1), mfrow = c(2, 2))
barplot(
  height = xrd_moreno_eig[1:10, 1],
  names.arg = rownames(xrd_moreno_eig)[1:10],
  ylab = "Inertia (%)", las = 1
)

## Plot cumulative inertia
plot(
  x = 1:10,
  y = xrd_moreno_eig[1:10, 3],
  type = "b",
  pch = 16,
  col = "red",
  xaxt = "n",
  ylab = "Cumulative inertia (%)", las = 1
)
axis(side = 1, at = 1:10, labels = rownames(xrd_moreno_eig)[1:10])

## Active rows (F1 vs F2)
plot(
  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = "black"
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")

## Columns (F1 vs F2)
plot(
  x = xrd_moreno_col_coord[, 1],
  y = xrd_moreno_col_coord[, 2],
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = adjustcolor("black", alpha = 0.2)
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
```

```{r}
#| label: fig-plot-contrib
#| fig-cap: Columns contribution to the construction of the first two components. The thirty variables contributing the most to each component are marked by red lines. 
#| fig-width: 10
#| fig-height: 5
par(mar = c(0, 4, 1, 1), mfrow = c(2, 1), oma = c(3, 0, 0, 0))
plot(
  x = theta_scale,
  y = xrd_moreno_col_contrib[, 1],
  type = "l",
  las = 1,
  ylab = "Contribution to F1",
  xaxt = "n"
)
abline( #21.94, 27.76-27.9, 29.42
  v = theta_scale[order(xrd_moreno_col_contrib[, 1], decreasing = TRUE)[1:30]],
  col = adjustcolor(col = "red", alpha = 0.1)
)
# alkahest::peaks_find(
#   x = theta_scale, 
#   y = xrd_moreno_col_contrib[, 1], 
#   SNR = 2, 
#   m = 5
# )

plot(
  x = theta_scale,
  y = xrd_moreno_col_contrib[, 2],
  type = "l",
  las = 1,
  ylab = "Contribution to F2"
)
abline( #12.32, 26.66, 31.36
  v = theta_scale[order(xrd_moreno_col_contrib[, 2], decreasing = TRUE)[1:30]],
  col = adjustcolor(col = "red", alpha = 0.1)
)
# alkahest::peaks_find(
#   x = theta_scale,
#   y = xrd_moreno_col_contrib[, 2],
#   SNR = 2,
#   m = 5
# )
mtext(expression(2*theta), side = 1, line = 2, outer = TRUE)
```

```{r}
#| label: fig-plot-ca-loi
#| fig-cap: Row (left) and columns (right) coordinates along the first two CA components. For the sake of readability, only the three main peaks of a selection of minerals of interest are displayed.
#| fig-width: 10
#| fig-height: 10
layout(matrix(data = c(1, 2, 3, 3), ncol = 2))
par(mar = c(4, 4, 1, 1))

id_moreno <- match(rownames(xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, ]), corpus$sample)

## F1 vs F2: active only, by material
col_material <- khroma::color("high contrast")(2)
plot(
  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = col_material[as.factor(corpus$material[id_moreno])]
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
legend("bottomright", legend = levels(as.factor(corpus$material[id_moreno])),
       pch = 16, col = col_material)

## F1 vs F2: active only, by loi
loi_moreno <- corpus$LOI[id_moreno]
loi_NA <- is.na(loi_moreno)
loi_moreno[loi_NA] <- min(loi_moreno, na.rm = TRUE)
col_loi <- khroma::ramp(loi_moreno, palette = "YlOrBr")(5)
col_loi[loi_NA] <- "darkgrey"
attr(col_loi,"breaks")[loi_NA] <- NA
loi_moreno[loi_NA] <- NA

plot(
  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
  y = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 2],
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 21,
  las = 1, asp = 1,
  col = "black",
  bg = col_loi
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")

## F1 vs LOI: active only
data_fit <- data.frame(
  loi = loi_moreno,
  F1 = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1]
)
loi_fit <- lm(loi ~ poly(F1, 2), data = data_fit)
seq_F1 <- seq(from = min(data_fit$F1), to = max(data_fit$F1), by = 0.1)
loi_pred <- predict(
  object = loi_fit, 
  newdata = data.frame(F1 = seq_F1),
  interval = "confidence", 
  level = 0.95
)

plot(
  x = xrd_moreno_row_coord[!xrd_moreno_row_coord$.sup, 1],
  y = loi_moreno,
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = "LOI (%)",
  type = "p",
  pch = 21,
  las = 1,
  col = "black",
  bg = col_loi
)
polygon(
  x = c(seq_F1, rev(seq_F1)),
  y = c(loi_pred[, "lwr"], rev(loi_pred[, "upr"])),
  col = adjustcolor("black", alpha = 0.2),
  border = NA,
  lwd = 2
)
lines(
  x = seq_F1,
  y = loi_pred[, "fit"], 
  col = "red",
  lwd = 2
)

legend_image <- as.raster(sort(col_loi[col_loi != "darkgrey"]))
rasterImage(legend_image, 0.80, 15, 0.90, 20)
rect(0.80, 15, 0.90, 20)
text(
  x = 0.9, 
  y = c(15, 20), 
  labels = paste0(round(range(attr(col_loi, "breaks"), na.rm = TRUE)), "%"), 
  pos = 4
)
```

```{r}
#| label: fig-plot-xrd
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(3, 1), mar = c(1, 0, 1, 1), oma = c(4, 5, 0, 0))

plot(
  x = theta_scale,
  y = xrd_counts[rownames(xrd_counts) == "BDX14922", ] / 1000,
  type = "l",
  xaxt = "n",
  ylab = "",
  las = 1
)
text(x = min(theta_scale) + 1, y = 55, labels = "BDX14922", font = 2)
plot(
  x = theta_scale,
  y = xrd_counts[rownames(xrd_counts) == "BDX15036", ] / 1000,
  type = "l",
  xaxt = "n",
  ylab = "",
  las = 1
)
text(x = min(theta_scale) + 1, y = 35, labels = "BDX15036", font = 2)
plot(
  x = theta_scale,
  y = xrd_counts[rownames(xrd_counts) == "BDX22203", ] / 1000,
  type = "l",
  ylab = "",
  las = 1
)
text(x = min(theta_scale) + 1, y = 15.5, labels = "BDX22203", font = 2)

mtext(expression(2*theta), side = 1, line = 3, cex = 1, outer = TRUE)
mtext(expression("Count ("*10^3*")"), side = 2, line = 3, cex = 1, outer = TRUE)
```

```{r}
#| label: ca-ceramics
## Ceramics
names_ok <- xrd_moreno_row_coord |> 
  subset(F2 < 0) |> 
  rownames()

# id_OK <- match(names_ok, rownames(xrd_counts))
id_ceram <- match(corpus$sample[corpus$material != "clay"], rownames(xrd_counts))
xrd_ceram <- xrd_counts[id_ceram, ]
sup_row_id <- match(corpus$sample[corpus$site != "Mas de Moreno"], rownames(xrd_ceram))
xrd_ceram_ca <- dimensio::ca(object = xrd_ceram, sup_row = sup_row_id)

## Get CA results
xrd_ceram_eig <- dimensio::get_eigenvalues(xrd_ceram_ca)
xrd_ceram_row_coord <- dimensio::get_coordinates(xrd_ceram_ca, margin = 1)
xrd_ceram_col_coord <- dimensio::get_coordinates(xrd_ceram_ca, margin = 2)
xrd_ceram_col_contrib <- dimensio::get_contributions(xrd_ceram_ca, margin = 2)
```

```{r}
#| label: fig-plot-minerals
#| fig-cap: XXX.
#| fig-width: 7
#| fig-height: 7
theta_minerals <- vapply(
  X = minerals$theta,
  FUN = function(x, theta) { which.min(abs(theta - x)) },
  FUN.VALUE = numeric(1),
  theta = theta_scale
)
xrd_minerals_coord <- xrd_ceram_col_coord[theta_minerals, ]
xlim <- range(range(xrd_ceram_row_coord[, 1]), range(xrd_minerals_coord[, 1]))
ylim <- range(range(xrd_ceram_row_coord[, 2]), range(xrd_minerals_coord[, 2]))

## Columns (F1 vs F2)
par(mar = c(4, 4, 1, 1))
col_minerals <- khroma::color("bright")(6)
plot(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  xlim = xlim, ylim = ylim,
  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = col_minerals[as.factor(minerals$symbole)]
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
text(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  labels = mapply(
    FUN = function(s, h, k, l) {
      if (h < 0) as.expression(bquote(.(s)~"("*bar(.(abs(h)))*.(k)*.(l)*")"))
      else as.expression(bquote(.(s)~"("*.(h)*.(k)*.(l)*")"))
    }, 
    s = minerals$symbole,
    h = minerals$h, 
    k = minerals$k,
    l = minerals$l
  ),
  col = col_minerals[as.factor(minerals$symbole)],
  pos = c(3, 2, 1, 1, 2, 4, 2, 2, 2, 4, 3, 3, 1, 3, 1, 4)
)
```

```{r}
#| label: fig-plot-ca-ceram
#| fig-cap: XXX.
#| fig-width: 10
#| fig-height: 10
layout(matrix(data = c(1, 2, 3, 3), ncol = 2))
par(mar = c(4, 4, 1, 1))

id_moreno <- match(rownames(xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, ]), corpus$sample)
id_settlements <- match(rownames(xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, ]), corpus$sample)

## F1 vs F2: active only, by site
col_site <- khroma::color("bright")(2)
plot(
  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
  y = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 2],
  xlim = xlim, ylim = ylim,
  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
  type = "p",
  pch = 21,
  las = 1, asp = 1,
  bg = "lightgrey",
  col = "black"
)
points(
  x = xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 1],
  y = xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 2],
  pch = 24,
  bg = col_site[as.factor(corpus$site[corpus$site != "Mas de Moreno"])]
)
# car::dataEllipse(
#   x = as.matrix(xrd_ceram_row_coord[xrd_ceram_row_coord$.sup, 1:2]),
#   groups = as.factor(corpus$site[id_settlements]),
#   level = 0.95,
#   col = col_site,
#   plot.points = FALSE,
#   center.pch = NULL,
#   add = TRUE
# )
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
legend("bottomright", legend = levels(as.factor(corpus$site[id_settlements])),
       pch = 16, col = col_site)

## F1 vs F2: active only, by loi
loi_moreno <- corpus$LOI[id_moreno]
loi_NA <- is.na(loi_moreno)
loi_moreno[loi_NA] <- min(loi_moreno, na.rm = TRUE)
col_loi <- khroma::ramp(loi_moreno, palette = "YlOrBr")(5)
col_loi[loi_NA] <- "darkgrey"
attr(col_loi,"breaks")[loi_NA] <- NA
loi_moreno[loi_NA] <- NA

plot(
  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
  y = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 2],
  xlim = xlim, ylim = ylim,
  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_ceram_eig[2, 2], 1)),
  type = "p",
  pch = 21,
  las = 1, asp = 1,
  col = "black",
  bg = col_loi
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")

## F1 vs LOI: active only
data_fit <- data.frame(
  loi = loi_moreno,
  F1 = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1]
)
loi_fit <- lm(loi ~ poly(F1, 2), data = data_fit)
seq_F1 <- seq(from = min(data_fit$F1), to = max(data_fit$F1), by = 0.1)
loi_pred <- predict(
  object = loi_fit, 
  newdata = data.frame(F1 = seq_F1),
  interval = "confidence", 
  level = 0.95
)

plot(
  x = xrd_ceram_row_coord[!xrd_ceram_row_coord$.sup, 1],
  y = loi_moreno,
  xlab = sprintf("F1 (%g%%)", round(xrd_ceram_eig[1, 2], 1)),
  ylab = "LOI (%)",
  type = "p",
  pch = 21,
  las = 1,
  col = "black",
  bg = col_loi
)
polygon(
  x = c(seq_F1, rev(seq_F1)),
  y = c(loi_pred[, "lwr"], rev(loi_pred[, "upr"])),
  col = adjustcolor("black", alpha = 0.2),
  border = NA,
  lwd = 2
)
lines(
  x = seq_F1,
  y = loi_pred[, "fit"], 
  col = "red",
  lwd = 2
)

legend_image <- as.raster(sort(col_loi[col_loi != "darkgrey"]))
rasterImage(legend_image, 0.80, 10, 0.90, 15)
rect(0.80, 10, 0.90, 15)
text(
  x = 0.9, 
  y = c(10, 15), 
  labels = paste0(round(range(attr(col_loi, "breaks"), na.rm = TRUE)), "%"), 
  pos = 4
)
```
